---
title: "The 7 (+7) World Wonders"
author: "Chiara Di Gravio"
date: "October 14, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The map shows the so-called World Wonders. 

The red dots show where the 7 Ancient World Wonders were. The green dots show where the 7 New World Wonders are. If you click on each dot a pop-up with the name of the corresponding Wonder will appear. By clicking on the name you are re-direct to a Wikipedia page with a bit of history.

As a preview, you might want to know that only one of the 7 Ancient World Wonders is still in existence (can you guess which one?), so I think there are 8 World Wonders!

```{r, echo=FALSE}
# create a map
library(leaflet)

# Ancient
pyramid <- c(29.97416777, 31.1339477975)
gardens <- c(32.478476, 44.441039)
zeus <- c(37.6379, 21.6300)
temple <- c(37.9498, 27.3639)
mausoleum <- c(37.0381, 27.4244)
rhodi <- c(36.4349631, 28.081913)
lighthouse <- c(31.15458, 29.84587)

dat <- rbind(pyramid, gardens, zeus, temple, mausoleum, rhodi, lighthouse)
colnames(dat) <- c("latitude", "longitude")

# New
jesus <- c(-22.9519, -43.2105)
tajmahal <- c(27.1750, 78.0422)
MachuPicchu <- c(-13.1631, -72.5450)
ChichenItza <- c(20.6843, -88.5678)
Colosseum <- c(41.8902, 12.4922)
Petra <- c(30.3285, 35.4444)
GreatWall <- c(40.4319, 116.5704)

dat <- rbind(dat,jesus,tajmahal,MachuPicchu,ChichenItza,Colosseum, Petra, GreatWall)

dat <- data.frame(dat, Time = c(rep("Ancient",7),rep("New",7)))

pal <- colorFactor(c("red", "green"), domain = c("Ancient", "New"))

# add name to popup plus links
Wonders <- c(
      "<a href='http://en.wikipedia.org/wiki/Great_Pyramid_of_Giza'>Great Pyramid of Giza</a>",
      "<a href='http://en.wikipedia.org/wiki/Hanging_Gardens_of_Babylon'>Hanging Gardens of Babylon</a>",
      "<a href='http://en.wikipedia.org/wiki/Temple_of_Artemis'>Temple of Artemis</a>",
      "<a href='http://en.wikipedia.org/wiki/Statue_of_Zeus_at_Olympia'>Statue of Zeus at Olympia</a>",
      "<a href='http://en.wikipedia.org/wiki/Mausoleum_at_Halicarnassus'>Mausoleum at Halicarnassus</a>",
      "<a href='http://en.wikipedia.org/wiki/Colossus_of_Rhodes'>Colossus of Rhodes</a>",
      "<a href='http://en.wikipedia.org/wiki/Lighthouse_of_Alexandria'>Lighthouse of Alexandria</a>",
      "<a href='http://en.wikipedia.org/wiki/Christ_the_Redeemer_(statue)'>Christ the Redeemer</a>",
      "<a href='http://en.wikipedia.org/wiki/Taj_Mahal'>Taj Mahal</a>",
      "<a href='http://en.wikipedia.org/wiki/Machu_Picchu'>Machu Picchu</a>",
      "<a href='http://en.wikipedia.org/wiki/Chichen_Itza'>Chichen Itza</a>",
      "<a href='http://en.wikipedia.org/wiki/Colosseum'>The Colosseum</a>",
      "<a href='http://en.wikipedia.org/wiki/Petra'>Petra</a>",
      "<a href='http://en.wikipedia.org/wiki/Great_Wall_of_China'>Great Wall of China</a>"
)

dat %>% leaflet() %>% addTiles() %>% 
      addCircleMarkers(lng = ~longitude, lat = ~latitude, popup = Wonders,
                       color = ~pal(Time),
                       stroke = FALSE, fillOpacity = 1,
                       radius = ~4)
```