---
title: 'Assignment 3: Hospital Quality'
author: "Chiara Di Gravio"
date: "27 maggio 2016"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
```

## Assignment 3: Hospital Quality

The [Hospital Compare web site](http://hospitalcompare.hhs.gov), run by the U.S. Department of Health and Human Services, provides data and information about the quality of care at over 4,000 Medicare-certified hospitals in the U.S. Here we focused on three outcomes:

- heart attack

- heart failure

- pneumonia

and we look at which hospital has the lower death rate. The function needed for this assignment can be found in [ProgrammingAssignment3.R](https://github.com/ChiaraDG/datasciencecoursera/blob/master/R%20programming/ProgrammingAssignment3/ProgrammingAssignment3.R). First we look at the hospital that has the best (i.e. lowest) 30-day mortality for one of the three outcomes listed above in a pre-specified State.

```{r CallingFunction}
source("ProgrammingAssignment3.R")
```

For instance if we are interested in Texas:

```{r bestTexas}
best("TX","heart attack")
best("TX","heart failure")
best("TX", "pneumonia")
```

or if we want to look at the best hospitals in New Jersery:

```{r bestNJ}
best("NJ","heart attack")
best("NJ","heart failure")
best("NJ", "pneumonia")
```

Second, we rank the hospitals by outcome in each state. For instance, the 5 hospitals in Texas with lower 30-day mortality rate for pneumonia are:

```{r bestTexasPneumonia,warning=FALSE}
best5 <- rankhospital("TX", "heart failure", 1:5)
data.frame(best = best5)
```

whereas the worst hospital in Texas for pneumonia (i.e. hospital with highest 30-day mortality rate) is:

```{r worstTexasPneumonia,warning=FALSE}
rankhospital("TX", "pneumonia", "worst")
```

Finally, we look at which hospital in each State has a pre-specified ranking for the outcome of interest. For instance, the three hospitals that have the highest pneumonia 30-day mortality rate within their State are:

```{r}
tail(rankall("pneumonia","worst"),3)
```

whereas three hospitals that have the lowest pneumonia 30-day mortality rate within their Statse are:

```{r}
head(rankall("pneumonia","best"),3)
```